---
output: github_document
bibliography: inst/REFERENCES.bib
csl: readme.csl
nocite: '@*'
link-citations: true
---

<!-- README.md is generated from readme.Rmd. Please edit that file -->

```{r init_knitr, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  fig.width = 5,
  fig.height = 3.5,
  message = FALSE
)

# Hide warnings when knitting.
options(warn = -1)
```

# varimpact - variable importance through causal inference

[![R-CMD-check](https://github.com/ck37/varimpact/actions/workflows/check-standard.yaml/badge.svg)](https://github.com/ck37/varimpact/actions/workflows/check-standard.yaml)
[![codecov](https://codecov.io/gh/ck37/varimpact/branch/master/graph/badge.svg)](https://codecov.io/gh/ck37/varimpact)

## Summary

varimpact uses causal inference statistics to generate variable importance estimates for a given dataset and outcome. It answers the question: which of my Xs are most related to my Y? Each variable's influence on the outcome is estimated semiparametrically, without assuming a linear relationship or other functional form, and the covariate list is ranked by order of importance. This can be used for exploratory data analysis, for dimensionality reduction, for experimental design (e.g. to determine blocking and re-randomization), to reduce variance in an estimation procedure, etc. See @hubbard2018dataadaptive for more details, or @hubbard2016mining for an earlier description.

## Details

Each covariate is analyzed using targeted minimum loss-based estimation ([TMLE](https://CRAN.R-project.org/package=tmle)) as though it were a treatment, with all other variables serving as adjustment variables via [SuperLearner](https://github.com/ecpolley/SuperLearner). Then the statistical significance of the estimated treatment effect for each covariate determines the variable importance ranking. This formulation allows the asymptotics of TMLE to provide valid standard errors and p-values, unlike other variable importance algorithms.

The results provide raw p-values as well as p-values adjusted for false discovery rate using the Benjamini-Hochberg procedure [@benjamini1995controlling]. Adjustment variables are automatically clustered hierarchically using HOPACH [@vanderlaan2003hopach] in order to reduce dimensionality.  The package supports multi-core and multi-node parallelization, which are detected and used automatically when a parallel backend is registered. Missing values are automatically imputed using K-nearest neighbors [@troyanskaya2001missing; @jerez2010missing] and missingness indicator variables are incorporated into the analysis.

varimpact is under active development so please submit any bug reports or feature requests to the [issue queue](https://github.com/ck37/varimpact/issues), or email Alan and/or Chris directly.

## Installation

### GitHub

```{r install, eval=FALSE}
# Install remotes if necessary:
# install.packages("remotes")
remotes::install_github("ck37/varimpact")
```

### CRAN

varimpact is not on CRAN yet; install from GitHub as above.

## Examples

### Example: basic functionality

```{r example_1, fig.width = 6.5, fig.height = 3.5, fig.alt = "Horizontal bar chart titled Impact of V2: the adjusted outcome mean for each of V2's two bins, plus a third bar for the impact estimate. Bars are colored to mark the lower-risk level, the higher-risk level and the impact."}
library(varimpact)

####################################
# Create test dataset.
set.seed(1, "L'Ecuyer-CMRG")
N <- 200
num_normal <- 4
X <- as.data.frame(matrix(rnorm(N * num_normal), N, num_normal))
Y <- rbinom(N, 1, plogis(.2*X[, 1] + .1*X[, 2] - .2*X[, 3] + .1*X[, 3]*X[, 4] - .2*abs(X[, 4])))
# Add some missing data to X so we can test imputation.
for (i in 1:10) X[sample(nrow(X), 1), sample(ncol(X), 1)] <- NA

####################################
# Basic example
vim <- varimpact(Y = Y, data = X)

# Review consistent and significant results.
vim

# Look at all results.
vim$results_all

# Plot the V2 impact.
plot_var("V2", vim)

# Generate latex tables with results.
exportLatex(vim)

# Clean up LaTeX files
cleanup_latex_files()
```

### Example: customize outcome and propensity score estimation

```{r example_2}
Q_lib = c("SL.mean", "SL.glmnet", "SL.ranger", "SL.rpartPrune")
g_lib = c("SL.mean", "SL.glmnet")
set.seed(1, "L'Ecuyer-CMRG")
(vim = varimpact(Y = Y, data = X, Q.library = Q_lib, g.library = g_lib))
```

### Example: parallel via multicore

```{r example_3}
library(future)
plan("multisession")
vim = varimpact(Y = Y, data = X)
```

### Example: mlbench breast cancer

```{r example_5, fig.width = 6, fig.height = 3.5, fig.alt = "Horizontal bar chart titled Impact of Mitoses: the adjusted outcome mean for Mitoses levels 1, 2 and 3, which increases with level, plus a fourth bar for the impact estimate. Bars are colored to mark the lower-risk level, intermediate levels, the higher-risk level and the impact."}
data(BreastCancer, package = "mlbench")
data = BreastCancer

# Create a numeric outcome variable.
data$Y = as.integer(data$Class == "malignant")

# Use multicore parallelization to speed up processing.
plan("multisession")
(vim = varimpact(Y = data$Y, data = subset(data, select = -c(Y, Class, Id))))
plot_var("Mitoses", vim)
```

## Authors

Alan E. Hubbard and Chris J. Kennedy, University of California, Berkeley

## References

::: {#refs}
:::
